Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging

### 2023-01-08 script/manuscript_MEE.Rmd
stable going to try lua for author formatting

---
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging
  Patterns of Queen Bumble Bees
author: |
  Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie $^3$, Emily J. Woodworth $^1$$^,$$^2$, Sophie Taddeo $^1$$^,$$^2$, Paul J. CaraDonna $^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant $^1$$^,$$^2$ | $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022 USA,  $^2$Plant Biology and Conservation, Northwestern University, Evanston, IL 60208, USA, $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, CO 81224, USA
abstract: |
  .    
  1) DNA Barcoding has been remarkably successful in nearly all kingdoms of life and has allowed for the rapid analysis of ecological assemblies. Successful DNA barcoding in the plant kingdom has been more difficult than other kingdoms. Due to this understanding plants in ecological contexts and understandings of their synecology in some instances may begin to lag behind other kingdoms.
  2) Here we utilize hyb-seq, museum studies, and species distribution modelling, to detect the plant species present in pollen loads collected from Queen Bumble Bees.
  3) By utilizing Species distribution modelling we allow for one to process hyb-seq data, create user specified sequence databases which may use MORE ACCURATE alignment algorithms on personal computers over realistic time periods. 
  4) We show that hyb-seq using the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches in real world scenarios.
  5) We conclude that these probes offer promise for the identification of plant tissue in both single sample, and metasample contexts.

output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: true
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

  return(writeLines(paste0('were detected in ',  detected, 
                '% of samples where the proportion of target grains were between ', 
                Lprop, '-', Hprop,  ' (n = ', total, ' Mdn = ', Mdn, ')')))
}
```

@@

*< UTILITY OF MASS IDENTIFICATION OF MATERIAL - FRIENDLY STAGE SET >*

*< ISSUES AND COSTS WITH IDENTIFYING MATERIAL TO SPECIES >* many spp defined by interactions (<;

The inability to reliably identify plants down to species can limit our understanding of ecosystem function and interactions. This is especially true for genera where the species are well defined based upon ecological and behavioral rather than morphological properties, and hence can serve as key habitat bioindicators(e.g. different species of Sagebrush- *Artemisia* L., Willows - *Salix* L., and Sedges - *Carex* L.) (@Gage2013HistoricalRO). The lack of species level data can hinder our understanding of the breadth of habitat which some species occupy, and the interactions they have with other species (CITE). Current methods to ameliorate this situation include: ignoring these ecologically relevant levels of detail, revisiting plots as diagnostic material becomes temporally available, assistance from taxonomic specialists, or the use of barcoding or other molecular techniques.

@@

  *< BARCODING HAS ITS LIMITS CURRENTLY BUT THERE IS POTENTIAL >*

  Recently barcoding (the identification of a sample from a single organism *e.g.* a piece of leaf), and metabarcoding (the identification of a sample containing a mix of organisms *e.g.* soil), have shown considerable promise in all Kingdoms of life (@ruppert2019past). With plants the identification of members of certain clades using barcoding has been quite successful (REF), whereas with other clades results have been elusive (@liu2014identification, @china2011comparative, @coissac2012bioinformatic), however metabarcoding incurs additional challenges to those which  exist for the currently implemented barcodes (@li2015plant, @kress2007two, @cbol2009dna, @coissac2012bioinformatic). Particular challenges with barcodes include the utilization of high-copy number sequences are associated with their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes, @fazekas2009plant). Particular challenges with the utilization of high-copy number sequences are associated with their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes).

  *<  WHAT NOVEL APPROACES IN ADDITION TO A353 ?>*

  Currently the largest plant systematic endeavor ever undertaken, the Kew Plant and Fungal Tree of Life (PAFTOL), is approaching completion (@baker2021PAFTOL). This data set will contain Hyb-Seq data from at least one species representing each genus in the plant kingdom using the popular A353 probes (@baker2021PAFTOL), resulting in over 14,000 represented species. These publicly available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding. Data from the 10kP project, which seeks to develop reference genomes from a phylogenetically diverse suite of plants will contribute many more records upon it’s intended completion, now slated to be by 2030, similar projects which seek to sequence high amounts of genomes in regions e.g. the ‘Darwin Tree of Life’ are being undertaken which will contribute data  applicable to enormous spatial domains (@cheng2018tenkp, @darwin2022project, @lewin2022biogenome). These data will promote the ability to apply metabarcoding to resolve a diversity of questions relevant to theoretical and applied ecology (cite). However, the application of metabarcoding still face challenges relating to the enormity of the genomic data sets and the computational power required to process sequence data. Herein we have resolved major components of the problems of identifying plant material without diagnostic morphological character states using the Angiosperms353 (A353) Hyb-Seq probes (@johnson2019universal), and custom species sequence databases derived via species distribution modelling, and temporal filtering.

  *< THIS P -- WHAT WE DO TO TEST THIS OUT >*

  To increase the quality of metabarcoding results in plants, we suggest reducing the number of possible plant species candidates by generating user selected sequence databases relevant to the the study region and its ecological characteristics (CITE !?). To achieve this goal, we first create a list of candidate species using digital collections gleaned from herbaria, survey work, and citizen science (e.g. iNaturalist), from a region exceeding the study area. To these candidate species, modelling approaches - such as logistic regression, may be used to identify taxa which warrant further exploration e.g. modelling to determine their possibility of presence in metabarcoding samples. We then use species distribution models to create potential distribution maps for the candidate species to limit the impact of spatial and taxonomic biases in the species list and account for spatial variations in niche availability throughout the study area. Species distribution models (SDM’s) examine the ecological conditions associated with known occurrence of a species to identify where else in the study area might suitable habitats be found. This approach has the additional benefit of greatly reducing the size of a sequence database, which allows for the usage of genomic size data on personal computers. This can also significantly reduce processing time, particularly as as most next-generation sequence data is deposited as raw-sequence reads.

@@

  To test these metagenomic and informatics approaches to determine whether the foraging record of Queen Bumble Bee's is consistent across direct observations and the pollen record, an incongruency in several floral visitation networks involving smaller bodied fauna (@barker2021pollen, @zhao2019topology, @alarcon2010congruence). The assessment of the plant species compositions of pollen is a desired results, with several applications, and numerous complications (@pornon2017dna, @bell2017applying, @sickel2015increased, @bell2019quantitative, @suchan2019pollen,  @johnson2021airborne). The two foraging phases of the Queen Bumble Bee life cycle is essential to 1) increase their weight before diapause, 2) increase their ovary weights while establishing their recently found nests, both of these time periods represent potential demographic bottlenecks in bumble bee populations (@sarro2022bumble). Bumblebees are one of the only groups of insects with unequivocal quantitative evidence for numerous populations declines, while simultaneously serving as the most effective pollinators in temperate montane ecosystems (@cameron2020global, @goulson2008decline, @williams1982distribution, @colla2012assessing, @bergman1996micrometeorological, @bingham1998efficient). *Heterogeneous montane ecosystems represent one of the most ecologically resilient areas in the temperate and oftentimes offer the sole potential refugia for multiple dimensions of biodiversity under climate change, whilst simultaneously experiencing the most abrupt changes in mean annual temperature (CITE). An immediate understanding of how to manage previously overlooked keystone insect species, such as bumble bees, is essential if these refugial ecosystemn services are to be utilized while maintaining their current species richness.*

@@

Observations and sample collection was conducted at The Rocky Mountain Biological Laboratory (RMBL; 38°57.5” N, 106°59.3” W (WGS 84), 2900 m.a.s.l.), Colorado, USA (APPENDIX 1 for site information). Pollinator observations of Bombus Latreille spp. (Apidae Latreille) were conducted from June - August of 2015 in six study sites characterized by high-montane/subalpine Parkland vegetation communities. Observations of Bombus foraging took place for one hour at each field site in three 100m transects, where all flowers were also counted and placed into abundance bins. Corbiculae loads were, non-lethally, collected once from all Queen individuals encountered.

@@

In order to minimise the number of species for which SDM’s were to be generated, BIEN was queried at a distance of up to 100km from our study area and all plant species records were downloaded. **_In order to emulate the stochasticity of botanical collecting, this data set was bootstrap re-sampled 250 times, with 90% of samples selected, to create a testing data set_**. The median of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance, under which, to include species as candidates for distribution modelling.

@@

535 species were modelled using Generalized Linear Models and Generalized Additive Models. 534 species were modelled using Random Forest and Boosted Regression Trees.  To evaluate the accuracy of the species distribution models, additional presence records from GBIF (n = 61,789), and AIM (n = 12,730) were used as test and training sets (n = 74,519) for logistic regression (CITE AIM AND @gbifDL2021sdms). Additional novel absence records were generated from the AIM data set to create a data set where each species has balanced presence and absences. 11 or more paired presence and absence records were required for this testing, resulting in 334 species being included in the logistic regression (Mdn = 110.0, $\bar{x}$ = 223.1, max = 1568 record pairs used) with a 70% test split (@caret).

@@

Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites, we identified the plant taxa most frequently visited by Queens across all years. We sequenced the 12 most commonly visited taxa twice using samples from one site within the Gunnison River Drainage and one individual from another population. In addition, for any of these 12 focal species which did not have a congener pair in this filtered sample, we included a congener - or a species from a closely related genus to serve as an outgroup. We also sequenced another 15 abundant taxa commonly visited by *Bombus* workers, based on the aforementioned data set (*APPENDIX 4*).

@@

A local NCBI database was built using the same processed novel and downloaded sequences (@camacho2009blast).

@@

To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  We used Divisive Hierarchical Clustering techniques to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner. 
  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Cerca* 0.5mm^2 of pollen was placed onto a ~4mm^2 fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrants) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with nail polish; all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained an unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round of sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.

@@

```{r, echo = F}
#a <- c('Trifolium', 'Lupinus', 'Glyccrrhiza', 'Mitella', 'Geum')
#b <- c('Lupinus', 'Lathyrus', 'Potentilla', 'Androsace', 'Bistorta', 'Vicia')
#c <- c('Jeffersonia', 'Micranthes', 'Prunus', 'Delphinium', 'Androsace', 
#       'Penstemon', 'Orthocarpus', 'Scuttelaria', 'Aquilegia', 'Castilleja', 'Draba')
#d <- c('Salix', 'Boechera')

@@

# set up as evaluation table !?

@@

M1 <- glm(DETECTED ~ Prop, mertensia_detected, family = binomial)
M.df <- data.frame(Prop = seq(0, 1, 0.0001))
M.df$DETECTED <- predict(M1, newdata=M.df, type="response")

ggplot(mertensia_detected, aes(x=Prop, y=DETECTED)) +
  stat_smooth(method="glm", se=T, fullrange=TRUE, 
              method.args = list(family=binomial))  +
  theme_bw()

# Repeat for Asteraceae 1 & 2

blst_aster <- blst %>% 
  filter(Family == 'Asteraceae') %>% 
  select(sample.id,  Prcnt_seqs, Genus) %>% 
  group_by(sample.id) %>% 
  distinct(sample.id, .keep_all = T)

@@

M1 <- glm(DETECTED ~ Prop, aster2_detected, family = binomial)
M.df <- data.frame(Prop = seq(0, 1, 0.0001))
M.df$DETECTED <- predict(M1, newdata=M.df, type="response")

ggplot(aster2_detected, aes(x=Prop, y=DETECTED)) +
  stat_smooth(method="glm", se=T, fullrange=TRUE, method.args = list(family='binomial'))  +
  theme_bw()

@@

ggplot(aster1_detected, aes(x=Prop, y=DETECTED)) +
  stat_smooth(method="glm", se=T, fullrange=TRUE, method.args = list(family='binomial'))  +
  theme_bw()

@@

rm(M1, M.df, M.df,  mertensia_samples, blst, blst_aster, 
   blst_aster1, blst_aster2, aster1_samples,  aster2_samples, 
   a, b, c, d, blast_mertensia, aster_samples)
```

The pollen samples contained three morphotypes which could readily be identified. Two of these mapped to monophyletic clades (Boraginaceae & Heliantheae Alliance), and one to a paraphyletic clade (Asteraceae less Heliantheae).
Boraginaceae grains `r prcnt_detect(mertensia_detected, Lprop = 0.01, Hprop = 1)`.  Asteraceae type 1, non-helianthoids, `r prcnt_detect(aster1_detected, Lprop = 0.001, Hprop = 0.01)` Asteraceae type 2, Helianthoids, `r prcnt_detect(aster2_detected, Lprop = 0.001, Hprop = 0.01)`. Both morphotypes of Asteraceae pollen `r prcnt_detect(aster_detected, Lprop = 0.01, Hprop = 1)`.

@@

```{r}
rm(mertensia_detected, aster1_detected, aster2_detected, aster_detected)
```

@@

```{r clean sequence numbers from environment, eval = F, echo = F}
rm(toseq, krak, brack, blast, loci, no_loci)
```

@@

ggplot(data=correct, aes(x=Observation, y=value,
                         group = Bombus.Species, color = Bombus.Species)) +
  geom_line(lwd = 1.25)+
  geom_point(size = 3, shape = 23, aes(fill = Bombus.Species)) +
  scale_fill_manual('Bee species', values = bee_cols) +
  scale_color_manual('Bee species', values = bee_cols) +
  theme_bw() +
  scale_y_continuous(breaks = c(0,2,4,6,8,10), limits = c(0,10))+
  labs(title = 'Taxa shared between the top 10 observations and molecular work', 
       y = 'Shared Observations', x= NULL) + 
  theme(panel.grid.major.x = element_blank(),
        panel.border = element_blank())

@@

```{r}

@@

... get ...

@@

**overall stuff and things**
We have demonstrated how Angiosperms533 data may be used for plant barcoding in a metagenomic context. This was exemplified in an ecologically realistic situation, centered both in a context which serves as an immediate framework for land management studies, and which produced fundamental natural history information. These data sets included closely related taxa in notoriously taxonomically difficult clades (e.g. *Mertensia*, *Lupinus*). We utilized spatial and temporal approaches for creating custom sequence databases which are readily applicable to any lab group with the capacity to perform next-generation sequencing across the entirety of multiple continents. In conjunction with well designed and implemented field based data collection we show how these methods may be widely used to test a variety of hypotheses related to ecological interactions.

These results show that the overall results between **Bumble Bee ecology** observational and barcoding are largely congruent. But that ...
Future analyses of the long term data set... Global studies should aim to corroborate ...

**Spatial ... & Timing ** (filters!)
- Fewer modelling runs for SDM's likely to be effective for determining inclusion, elastic inclusion criteria per application.
- The actual data set which was used for training and testing all of the models incorporated into SDM's represented only roughly one quarter of the records available for such purposes. We consciously chose to do this in order to showcase the possibility of this approach working in less data rich areas. 
- Used fine scale flora to evaluate results of modelling, similar approaches based on quick field surveys which do not require comprehensive floristic endeavors.
- Timing in other areas may suffer complications due to lacking as strong a signal as that induced by snow melt, but longer duration of flowering likely to help. 
- Sample size in phenology estimates had an effect which is likely to be maintained across all results.
- Regardless both show good agreement on flower onset, peak flowering, and moderate agreement with flowering cessation. The disagreement in flowering cessation is perhaps due to more microclimates which retain water, rather than microclimates which allow the early accumulation of heat. 
- Number of botanists rapidly decreasing no hope for much more field based information or funding.

>>>>

---
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging
  Patterns of Queen Bumble Bees
author: |  
  Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie $^3$, Emily J. Woodworth $^1$$^,$$^2$, Sophie Taddeo $^1$$^,$$^2$, Paul J. CaraDonna $^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant $^1$$^,$$^2$ | $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022 USA,  $^2$Plant Biology and Conservation, Northwestern University, Evanston, IL 60208, USA, $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, CO 81224, USA]
abstract:  |
  1) DNA Barcoding has been remarkably successful in nearly all kingdoms of life and has allowed for the rapid analysis of ecological assemblages. Successful DNA barcoding in the plant kingdom has been more difficult than other kingdoms. Due to this understanding plants in ecological contexts and understandings of their synecology in some instances may begin to lag behind other kingdoms.  
  2) Here we utilize hyb-seq, museum studies, and species distribution modelling, to detect the plant species present in pollen loads collected from Queen Bumble Bees.  
  3) By utilizing Species distribution modelling we allow for one to process hyb-seq data, create user specified sequence databases which may use MORE ACCURATE alignment algorithms on personal computers over realistic time periods.  
  4) We show that hyb-seq using the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches in real world scenarios.  
  5) We conclude that these probes offer promise for the identification of plant tissue in both single sample, and metasample contexts.  
#keywords: | 
#"DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species #distribution model"
output:
  pdf_document: default
  toc: no
  pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: true
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

  r <- paste0('were detected in ',  detected, 
                '% of samples where the proportion of target grains were between ', 
                Lprop, '-', Hprop,  ' (n = ', total, ' Mdn = ', Mdn, ')')
  return(r)
}

@@

The inability to reliably identify plants down to species can limit our understanding of ecosystem function and interactions (@bortolus2008error). This is especially true for genera where species are defined based upon ecological and behavioral rather than morphological properties, and hence can serve as key habitat bioindicators (e.g. different species of Sagebrush- *Artemisia* L., Willows - *Salix* L., and Sedges - *Carex* L.) (@Gage2013HistoricalRO). The lack of species level data can hinder our understanding of the breadth of habitat which some species occupy, and the interactions they have with other species. Current methods to ameliorate this situation include: ignoring these ecologically relevant levels of detail, revisiting plots as diagnostic material becomes temporally available, assistance from taxonomic specialists, or the use of barcoding or other molecular techniques.

@@

  Recently barcoding (the identification of a sample from a single organism *e.g.* a piece of leaf), and metabarcoding (the identification of a sample containing a mix of organisms *e.g.* soil), have shown considerable promise in all Kingdoms of life (@ruppert2019past). With plants the identification of members of certain clades using barcoding has been quite successful (@kress2017plant), whereas with other clades results have been elusive (@liu2014identification, @china2011comparative, @coissac2012bioinformatic), however metabarcoding incurs additional challenges to those which  exist for the currently implemented barcodes (@li2015plant, @kress2007two, @cbol2009dna, @coissac2012bioinformatic). Particular challenges with barcodes include the utilization of high-copy number sequences are associated with their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes, @fazekas2009plant).

  Currently the largest plant systematic endeavor ever undertaken, the Kew Plant and Fungal Tree of Life (PAFTOL), is approaching completion (@baker2021PAFTOL). This data set will contain hybridization capture (Hyb-Seq) data from at least one species representing each genus in the plant kingdom using the popular A353 probes (@baker2021PAFTOL), resulting in over 14,000 represented species. These publicly available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding, and the A353 probes are currently being used in many other plant phylogenetic issues increasing the sampling depth of many clades (@baker2021exploring). Data from the 10kP project, which seeks to develop reference genomes from a phylogenetically diverse suite of plants will contribute many more records upon it’s intended completion, now slated to be by 2030, similar projects which seek to sequence high amounts of genomes in regions e.g. the ‘Darwin Tree of Life’ are being undertaken which will contribute data  applicable to enormous spatial domains (@cheng2018tenkp, @darwin2022project, @lewin2022biogenome). These data will promote the ability to apply metabarcoding to resolve a diversity of questions relevant to theoretical and applied ecology (@kress2017plant). However, the application of metabarcoding still face challenges relating to the enormity of the genomic data sets and the computational power required to process sequence data. Herein we have resolved major components of the problems of identifying plant material without diagnostic morphological character states using the Angiosperms353 (A353) Hyb-Seq probes (@johnson2019universal), and custom species sequence databases derived via species distribution modelling, and temporal filtering.

  To increase the quality of metabarcoding results in plants, we suggest reducing the number of possible plant species candidates by generating user selected sequence databases relevant to the the study region and its ecological characteristics. To achieve this goal, we first create a list of candidate species using digital collections gleaned from herbaria, survey work, and citizen science (e.g. iNaturalist), from a region exceeding the study area. To these candidate species, modelling approaches - such as logistic regression, may be used to identify taxa which warrant further exploration e.g. modelling to determine their possibility of presence in metabarcoding samples. We then use species distribution models to create potential distribution maps for the candidate species to limit the impact of spatial and taxonomic biases in the species list and account for spatial variations in niche availability throughout the study area. Species distribution models (SDM’s) examine the ecological conditions associated with known occurrence of a species to identify where else in the study area might suitable habitats be found. This approach has the additional benefit of greatly reducing the size of a sequence database, which allows for the usage of genomic size data on personal computers. This can also significantly reduce processing time, particularly as as most next-generation sequence data is deposited as raw-sequence reads.

@@

  To test these metagenomic and informatics approaches to determine whether the foraging record of Queen Bumble Bee's is consistent across direct observations and the pollen record, an incongruency in several floral visitation networks involving smaller bodied fauna (@barker2021pollen, @zhao2019topology, @alarcon2010congruence). The assessment of the plant species compositions of pollen is a desired results, with several applications, and numerous complications (@pornon2017dna, @bell2017applying, @sickel2015increased, @bell2019quantitative, @suchan2019pollen,  @johnson2021airborne). The two foraging phases of the Queen Bumble Bee life cycle is essential to 1) increase their weight before diapause, 2) increase their ovary weights while establishing their recently found nests, both of these time periods represent potential demographic bottlenecks in bumble bee populations (@sarro2022bumble). Bumblebees are one of the only groups of insects with unequivocal quantitative evidence for numerous populations declines, while simultaneously serving as the most effective pollinators in temperate montane ecosystems (@cameron2020global, @goulson2008decline, @williams1982distribution, @colla2012assessing, @bergman1996micrometeorological, @bingham1998efficient). Montane areas often represent the most diverse areas in the temperate and oftentimes offer the sole potential refugia for multiple dimensions of biodiversity under climate change, whilst simultaneously experiencing the greatest proportional changes in mean annual temperature (@brito2018climate, @pepin2022climate). An immediate understanding of how to manage previously overlooked keystone insect species, such as bumble bees, is essential if the refugial potential of the temperate mountains are to be utilized while maintaining their current diversity (@loarie2009velocity, @dobrowski2016climate)

@@

Observations and sample collection was conducted at The Rocky Mountain Biological Laboratory (RMBL; 38°57.5” N, 106°59.3” W (WGS 84), 2900 m.a.s.l.), Colorado, USA (APPENDIX 1 for site information). Pollinator observations of Bombus Latreille spp. (Apidae Latreille) were conducted from June - August of 2015 in six study sites characterized by high-montane/subalpine Parkland vegetation communities. Observations of *Bombus* foraging took place for one hour at each field site in three 100m transects, where all flowers were also counted and placed into abundance bins. Corbiculae loads were, non-lethally, collected once from all Queen individuals encountered.

@@

In order to minimise the number of species for which SDM’s were to be generated, BIEN was queried at a distance of up to 100km from our study area and all plant species records were downloaded. In order to emulate the stochasticity of botanical collecting, this data set was bootstrap re-sampled 250 times, with 90% of samples selected, to create a testing data set. The median of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance, under which, to include species as candidates for distribution modelling.

@@

535 species were modelled using Generalized Linear Models and Generalized Additive Models. 534 species were modelled using Random Forest and Boosted Regression Trees.  To evaluate the accuracy of the species distribution models, additional presence records from GBIF (n = 61,789), and AIM (n = 12,730) were used as test and training sets (n = 74,519) for logistic regression (@gbifDL2021sdms, @aim2019database). Additional novel absence records were generated from the AIM data set to create a data set where each species has balanced presence and absences. 11 or more paired presence and absence records were required for this testing, resulting in 334 species being included in the logistic regression (Mdn = 110.0, $\bar{x}$ = 223.1, max = 1568 record pairs used) with a 70% test split (@caret).

@@

Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites, we identified the plant taxa most frequently visited by Queens across all years. We sequenced the 12 most commonly visited taxa twice using samples from one site within the Gunnison River Drainage and one individual from another population. In addition, for any of these 12 focal species which did not have a congener pair in this filtered sample, we included a congener - or a species from a closely related genus to serve as an outgroup. We also sequenced another 15 abundant taxa commonly visited by *Bombus* workers, based on the aforementioned data set (*APPENDIX 4*). Plant collections were identified via a variety, and typically a combination, of dichotomous keys and primary literature as required (@flora1993flora, @hitchcock2018flora,  @ackerfield2015flora, @lesica2012manual,  @cronquist1977intermountain, @allred2012flora, @jepson2022online, @mohlenbrock2002vascular).

@@

A local NCBI database was built using the same processed novel and downloaded sequences as the previous databases (@camacho2009blast).

@@

To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  We used Divisive Hierarchical Clustering techniques to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner. 
  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Cerca* 0.5mm^2 of pollen was placed onto a ~4mm^2 fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrants) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with nail polish; all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained an unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round osf sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.

@@

```{r compare microscopic to molecular Pollen ID, echo = F}
a <- c('Trifolium', 'Lupinus', 'Glyccrrhiza', 'Mitella', 'Geum')
b <- c('Lupinus', 'Lathyrus', 'Potentilla', 'Androsace', 'Bistorta', 'Vicia')
c <- c('Jeffersonia', 'Micranthes', 'Prunus', 'Delphinium', 'Androsace', 
       'Penstemon', 'Orthocarpus', 'Scuttelaria', 'Aquilegia', 'Castilleja', 'Draba')
d <- c('Salix', 'Boechera')

@@

@@

blst_aster <- blst %>% 
  filter(Family == 'Asteraceae') %>% 
  group_by(sample.id) %>% 
  distinct(sample.id, .keep_all = T)

@@

@@

@@

aa <- prcnt_detect(mertensia_detected, Lprop = 0.01, Hprop = 1)
bb <- prcnt_detect(aster1_detected, Lprop = 0.001, Hprop = 0.01)
cc <- prcnt_detect(aster2_detected, Lprop = 0.001, Hprop = 0.01)
dd <-  prcnt_detect(aster_detected, Lprop = 0.01, Hprop = 1)

rm(mertensia_samples, blst, blst_aster, blst_aster1,
   blst_aster2, aster1_samples,  aster2_samples, a, b, c, d, 
   blst_mertensia, aster_samples)
```

@@

```{r, echo = F, warning = F, message = F}
rm(mertensia_detected, aster1_detected, aster2_detected, aster_detected)
```

@@

```{r clean sequence numbers from environment, eval = F, echo = F}
rm(toseq, krak, brack, blast, loci, no_loci, aa, bb, cc, dd)
```

To determine at which level species in pollen loads could be detected the results of light microscopy were compared to the molecular results. The pollen samples contained three morphotypes which could readily be identified via microscopy. Two of these mapped to the clades (Boraginaceae & Heliantheae Alliance), and one to a Asteraceae less Heliantheae. Boraginaceae grains `r toString(aa)`.  Asteraceae type 1, non-helianthoids, `r toString(bb)` Asteraceae type 2, Helianthoids, `r toString(cc)`. Both morphotypes of Asteraceae pollen `r toString(dd)`.

@@

ggplot(data=correct, aes(x=Observation, y=value,
                         group = Bombus.Species, color = Bombus.Species)) +
  geom_line(linewidth = 1.25)+
  geom_point(size = 3, shape = 23, aes(fill = Bombus.Species)) +
  scale_fill_manual('Bee species', values = bee_cols) +
  scale_color_manual('Bee species', values = bee_cols) +
  theme_bw() +
  scale_y_continuous(breaks = c(0,2,4,6,8,10), limits = c(0,10))+
  labs(title = 'Taxa shared between the top 10 observations and molecular work', 
       y = 'Shared Observations', x= NULL) + 
  theme(panel.grid.major.x = element_blank(),
        panel.border = element_blank())

@@

@@

@@

**overall stuff and things**
We have demonstrated how Angiosperms533 hyb-seq probes may be used for plant barcoding in a metagenomic context. This was exemplified in an ecologically realistic situation, centered both in a context which serves as an immediate framework for land management studies, and which produced fundamental natural history information. These data sets included closely related taxa in notoriously taxonomically difficult clades (e.g. *Mertensia*, *Lupinus*). We utilized spatial and temporal approaches for creating custom sequence databases which are readily applicable to any lab group with the capacity to perform next-generation sequencing across the entirety of multiple continents. In conjunction with well designed and implemented field based data collection we show how these methods may be widely used to test a variety of hypotheses related to ecological interactions.

*Complications* We likely had issues detecting pollen from several genera of plants, bsaed upon these and other obesrvationla studies most likely Vicia-Lathyrus and Frasera (CITE), probably due to low quality sequencing of whole plant material... (REED SHOULD HUNT FOR LOCI RETURNS FROM BLAST LIKE HE DID POLLEN)

These results show that the overall results between **Bumble Bee ecology** observational and barcoding are largely congruent. But that ...
Future analyses of the long term data set...
